A Realistic Guide to Making Data Available Alongside Code to Improve Reproducibility.

### 2019-12-19 paper.Rmd
relatively minor updates, including adding a new data codebook

Out of these sections, the minimal viable format are the following provided in one main directory folder:

@@

The README is often the first place people will go to learn more about data, and helps users navigate the repository. It is meant for someone to read and understand more about the data and contains the "who, what, when, where, why, & how":

@@

The README should be placed in the top level of the project. It should be brief, and provide links to the other aforementioned sections. Saving a README with the extension `.md` file gives the author the formatting benefits of `markdown`, making it simple to insert links, tables, and make lists. In systems like GitHub, a README file is detected and rendered in nice HTML on the repository by default.

@@

A data dictionary provide human readable description of the data, providing context on the nature and structure of the data. This helps someone not familiar with the data understand, and use the data. At a minimum they should contain:

@@

**Variable names** should be short, descriptive names with no spaces or special characters. In the most common tabular data case, these correspond to column names, for example, "job_position", "faculty_level", and "years_at_company". **Variable labels** are longer descriptions of variables. For example "University Job Position", "University Faculty Position Level", and  "Number of Years Worked at University" [@McGill-codebook; @Arslan2019]. **Variable codes** apply to categorical (factor) variables, and are the values for their contents. For example, a variable could contain answers to a question, with values and codes like so: `0 = no, 1 = yes`, or for another variable indicating indoors or outdoors: `0 = inside` and `1 = outside`. These should be consistent across similar variables to avoid problems where `0 = yes` for one variable, and `1 = yes` in another. Date variables should have consistent formatting. For example, all date information could be in format "YYYY-MM-DD", and this should not be mixed with "YYYY-DD-MM". Date formatting should be described in the variable labels. **Missing data** are values that should have been observed, but were not. The code for missingness should be documented in the data dictionary, and should nominally be `NA`. If the reason for missingness is known it should be recorded. For example, censored data, patient drop out, or measurement error can have different values, such as "unknown", -99, or other value codes [@White2013; @Broman2017].

@@

```{r table-data-incarcerate}
dictionary_incarcerate <- data.frame(
  Variable = c("year", "urbanicity", "pop_category", "rate_per_100000"),
  Class  = c("integer (date)",
             "character",
             "character",
             "double"),
  description = c("Year",
                  "County-type (urban, suburban, small/mid, rural)",
                  "Category for population - either race, gender, or Total",
                  "Rate within a category for prison population per 100,000 people")
)

knitr::kable(dictionary_incarcerate,
             caption = "The prisoner summary data dictionary, with columns on the variable, it's class, and a short description of the contents of the variable.",
             booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position"))
```

Data dictionary tables should be placed in the README and presented as a table. Every data dictionary should also be provided in its raw form in the repository, for example, saved as a CSV.

@@

The CC0 is a “public domain” license. Data with a CC0 license means the data owners waive all their rights to the work, and it now "owned" by the public. The data can be freely shared, this means it can be copied, modified, and distributed, even for commercial purposes _without asking permission_. When using data with CC0, it is good practice to cite the original paper, but it is not required. If you wish to use the CC0, see @choose-cc0.  For a brief overview of the CC0, see @cc0-short, and for the full license, see @cc0-long.

@@

A Digital Object Identifier (DOI), is a unique identifier for a digital object such as a paper, poster, or software, and is a prerequisite for citation. For example, if the data are deposited in dryad or Zenodo, the best practice would be to copy the citation created by these repositories. Under the hood, the DOI is "minted" by DataCite, who provide DOIs for research outputs and data. This means that when citing data, it only makes sense to cite datasets that have been deposited into a DataCite compliant repository. If a DOI is unavailable, a citation will be meaningless, as it cannot be tracked by any means.

@@

The README, and data dictionary provides _human readable_ information on the data. To ensure data types are preserved - dates are dates, names are characters, and so on - there needs to be some form of _machine readable_ metadata. Machine readable metadata also provide a structure that means the data can be indexed and searched. An excellent standard for metadata is Table Schema written by frictionless data [@fritionless-table-schema]. For example, a dataset called "demographics" with three variables is shown in Table \@ref(tab:table-machine-readable), which has the Java Script Object Notation (JSON) equivalent in Figure \@ref(fig:json-code).

@@

Another rich format is XML, the e**X**tensible **M**arkup **L**anguage. This was an early iteration on the idea of a plain text format that was both human and machine readable. It is very powerful - it powers the entire Microsoft Office suite, but for metadata purposes it is not as human readable as JSON. JSON is shorter, and quicker and easier to read and write than XML, and can also be parsed with standard JavaScript, whereas XML must be parsed by a special XML parser. XML and JSON can be converted into each others respective formats, for example with the `json` or `xml2` R packages. XML is still used as metadata for data storage, for example in EML, the Ecological Metadata Language [@EML-about].

@@

Unlike CRAN, bioconductor, does not have a 5 megabyte limit for package size and offers a more liberal data inclusion policy. They have a strict specification for how to organise genomic data in a package, so they can be used for data analysis out of the box. This is a robust solution, but only works for homogeneous data as found with bioinformatics. For example, data in their ExperimentData section (bioconductor.org/packages/3.9/data/experiment) can be used in a reliable way. However, such a strict standard would be impossible to enforce or scale on a repository as heterogeneous as CRAN.

However, these packages are still generally used for demonstration or testing purposes, over generating new knowledge for papers.

@@

There are a three common options for publishing data in research:

@@

In **Least Moving Parts**, the data might be published with an R package, or as part of a GitHub release using piggyback @piggyback, or a in serverless databases. This approach **gets the data somewhere rather than nowhere**. Its minimal features means it is simple to maintain. In addition, piggyback also keeps the data in the same location as the code, making it easier and simpler to organise. A downsides is that it does not scale to larger data. Self hosting the data is an option, but we discourage this, as it may succumb to bit rot - where data is corrupted, degraded, or servers turn off.

@@

Data used in publications are often shared in the supplementary materials of articles, or served on repositories such as the Dryad Digital Repository (@dryad). Dryad makes data from scientific publications discoverable, reusable, and citable. It is well funded through grants from the NSF, European Commission.

To provide better context around the data used in research and better expose data for reuse, journals are now adding "data papers". These are specifically designed for publishing articles about the data, and sharing it. This benefits both researchers and readers. Researchers receive credit for data they have collected or produced. Readers get more context about the data.

@@

Whilst useful, these requirements do not tell the author how to actually structure the data and folders. Instead providing ideas on what they should include. This is a useful step towards improving data reuse, but it _lacks some minimal structure that allows a researcher to have a predictable way to access and interpret the data_.

@@

The data are are free to use and publicly accessible, with the interface to the data being designed to cover a wide range of skills. For example, the marvin service to streamline access and visualisation of MaNGA data [@marvin-service], through to raw data access (see Figure \@ref(fig:sdss-raw-data-screenshot)).

@@

- Metadata in the machine readable standard, SCHEMA
- Data in multiple formats (e.g., .csv and .fits, database)
- Previously released data available in archives
- Entire datasets available for download
- Raw data before processed into database is also available

@@

1. **Decide whether publishing is appropriate for your data** 
It is critical to publish your data if they are the basis for a peer-reviewed research publication. To be broadly useful, your data must be deposited in a permanent archive, to ensure that it does not disappear from a ephemeral location such as your university website. It should also contain useful metadata so that someone outside the project team can understand what the data mean. However, this level of effort is not always critical for all data science efforts. For small-scale projects such as ones where one might generate simulated datasets, this level of curation is highly unnecessary. Here making data available in a transient location like GitHub or a website is sufficient.

@@

10. **Stick with simple data formats** to ensure long-term usefulness of datasets in a way that is not tied to transient technologies (such as ever changing spreadsheet standards), store the data as plain text (e.g., CSV) where possible. This can take the form of comma/tab separated files in most cases. This will ensure transparency and future proof your data.

@@

Over the years researchers have made data available in a multitude of unreliable ways, including offering downloads from university websites, FTP repositories, and password protected servers, all of which are prone to bit rot. In rare cases where the data are deposited in a permanent archive, insufficient metadata and missing data processing scripts have made it harder and time intensive to map these to code.  Our goal here is to describe a variety of ways in which a researcher can ship data, analogous to how easy it has become to push code to services like GitHub. The choice of technology and method all depends on the nature of the data, and its size and complexity. Once the issue of access is solved, we also discuss how much metadata to include in order to make the data useful. Ultimately there is a tradeoff to consider on a case by case basis. Long-term studies with complex protocols require structured metadata, while more standardized datasets such as those coming from sensors require far less effort.

>>>>

Out of these sections, the minimal files to provide, in order of most importance should be provided in one main directory folder:

@@

In the context of datasets, READMEs are particularly useful when there are no reliable standards. The best practice here is to have one README per dataset.

A README is often the first place people will go to learn more about anything in a folder - they are very common in softwar, and historically were includd so that the uppecase lettrs of README meant ti would be at the top of a directory. In the context of data, a README when there are no reliable standards. It is meant for someone to read and understand more about the data and contains the "who, what, when, where, why, & how":

@@

The README should be placed in the top level of the project, and with one README per dataset. It should be brief, and provide links to the other aforementioned sections, and be in one directory. Saving a README with the extension `.md` file gives the author the formatting benefits of `markdown`, making it simple to insert links, tables, and make lists. In systems like GitHub, a README file is detected and rendered in nice HTML on the repository by default.

@@

A data dictionary provide human readable description of the data, providing context on the nature and structure of the data. This helps someone not familiar with the data understand, and use the data. At a minimum they should contain the following pieces of information about the data:

@@

Figure \@ref(fig:variables) shows an example of the data and data dictionary details.
**Variable names** should be short, descriptive names with no spaces or special characters. In the most common tabular data case, these correspond to column names, for example, "job_position", "faculty_level", and "years_at_uni". **Variable labels** are longer descriptions of variables. For example "University Job Position", "University Faculty Position Level", and  "Number of Years Worked at University" [@McGill-codebook; @Arslan2019]. **Variable codes** apply to categorical (factor) variables, and are the values for their contents. For example, a variable could contain answers to a question, with values and codes like so: `0 = no, 1 = yes`, to indicate if someone is in statistics, for example.  These should be consistent across similar variables to avoid problems where `0 = yes` for one variable, and `1 = yes` in another. Date variables should have consistent formatting. For example, all date information could be in format "YYYY-MM-DD", and this should not be mixed with "YYYY-DD-MM". Date formatting should be described in the variable labels. **Missing data** are values that should have been observed, but were not. The code for missingness should be documented in the data dictionary, and should nominally be `NA`. If the reason for missingness is known it should be recorded. For example, censored data, patient drop out, or measurement error can have different values, such as "unknown", -99, or other value codes [@White2013; @Broman2017].

```{r variables, fig.cap = "Four tables - the data, and it's variable names, variable codes, and the meaning of missing values", fig.height = 8}
library(tibble)

dat <- tibble::tribble(
  ~job_position, ~outside, ~statistics, ~date_started,
   "Lecturer",        1,        1,      "2019-01-01",
   "Professor",       0,        0,      NA,
   "Senior Lecturer", 1,        1,      "unknown"
)

variable_labels <- tibble::tribble(
  ~variable, ~description,
  "job_position", "University Job Position",
  "outside",      "Did this person study outside this university?",
  "statistics",   "Is this perosn in the statistics faculty",
  "date_started", "What date did this perosn start their position?"
)

variable_codes <- tibble::tribble(
  ~variable,        ~code,
   "job_position",   "text description of job position",
   "outside",       "0 = not studied outside, \n1 = has studied outside",
   "statistics",    "0 = is not in statistics dept., \n1 = is currently in statistics dept.",
   "date_started",   "date variable in format YYYY-MM-DD, \ne.g., 2001-01-01 is 1st January, 2001"
)

missing_values <- tibble::tribble(
  ~value, ~meaning,
  NA,     "Missing value",
  "unknown", "Missing value"
)

library(gridExtra)
grid.arrange(
  tableGrob(dat),
  tableGrob(variable_labels),
  tableGrob(variable_codes),
  tableGrob(missing_values),
  ncol = 1
)

@@

```{r table-data-incarcerate}
dictionary_incarcerate <- data.frame(
  Variable = c("year", "urbanicity", "pop_category", "rate_per_100000"),
  Class  = c("integer (date)",
             "character",
             "character",
             "double"),
  Description = c("Year",
                  "County-type (urban, suburban, small/mid, rural)",
                  "Category for population - either race, gender, or Total",
                  "Rate within a category for prison population per 100,000 people")
)

knitr::kable(dictionary_incarcerate,
             caption = "The prisoner summary data dictionary, with columns for the variable, its class, and a short description of the contents of the variable.",
             booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position"))
```

Data dictionary tables should be placed in the README and presented as a table. Every data dictionary should also be provided in its raw form in the repository, for example, saved as a CSV, so that they don't become trapped in a README.

@@

The CC0 is a “public domain” license. Data with a CC0 license means the data owners waive all their rights to the work, and it now "owned" by the public. The data can be freely shared, copied, modified, and distributed, even for commercial purposes _without asking permission_. When using data with CC0, it is good practice to cite the original paper, but it is not required. If you wish to use the CC0, see @choose-cc0.  For a brief overview of the CC0, see @cc0-short, and for the full license, see @cc0-long.

@@

A Digital Object Identifier (DOI), is a unique identifier for a digital object such as a paper, poster, or software, and is a prerequisite for citation. For example, if the data are deposited in repositories like Dryad or Zenodo, the best practice would be to copy the citation created by these repositories. Under the hood, the DOI is "minted" by DataCite, who provide DOIs for research outputs and data. This means that when citing data, it only makes sense to cite datasets that have been deposited into a DataCite compliant repository. If a DOI is unavailable, a citation will be meaningless, as it cannot be tracked by any means.

@@

The README and data dictionary provides _human readable_ information on the data. To ensure data types are preserved - dates are dates, names are characters, and so on - there needs to be some form of _machine readable_ metadata. Machine readable metadata also provide a structure that allow the data to be indexed and searched online, through services scuh as google datasets search. An excellent standard for metadata is Table Schema written by frictionless data [@fritionless-table-schema]. For example, a dataset called "demographics" with three variables is shown in Table \@ref(tab:table-machine-readable), which has the Java Script Object Notation (JSON) equivalent in Figure \@ref(fig:json-code).

@@

Another rich format is XML, the e**X**tensible **M**arkup **L**anguage. This was an early iteration on the idea of a plain text format that was both human and machine readable. It is powerful and extensible (it powers the entire Microsoft Office suite), but for metadata purposes it is not as human readable as JSON. JSON is shorter, and quicker and easier to read and write than XML, and can also be parsed with standard JavaScript, whereas XML must be parsed by a special XML parser. XML and JSON can be converted into each others respective formats, for example with the `json` or `xml2` R packages. XML is still used as metadata for data storage, for example in EML, the Ecological Metadata Language [@EML-about].

@@

Unlike CRAN, bioconductor does not have a 5 megabyte limit for package size and offers a more liberal data inclusion policy. They have a strict specification for how to organise genomic data in a package, so they can be used for data analysis out of the box. This is a robust solution, but only works for homogeneous data as found with bioinformatics. For example, data in their ExperimentData section (bioconductor.org/packages/3.9/data/experiment) can be used in a reliable way. However, such a strict standard would be impossible to enforce or scale on a repository as heterogeneous as CRAN. However, these packages are still generally used for demonstration or testing purposes, over generating new knowledge for papers.

@@

There are three common options for publishing data in research:

@@

In **Least Moving Parts**, the data might be published with an R package, or as part of a GitHub release using piggyback @piggyback, or a in serverless databases. This approach **gets the data somewhere rather than nowhere**. Its minimal features means it is simple to maintain. In addition, piggyback also keeps the data in the same location as the code, making it easier and simpler to organise. A downside is that it does not scale to larger data. Self hosting the data is an option, but we discourage this, as it may succumb to bit rot - where data is corrupted, degraded, or servers turn off.

@@

Data used in publications are often shared in the supplementary materials of articles, or served on repositories such as the Dryad Digital Repository (@dryad). Dryad makes data from scientific publications discoverable, reusable, and citable. It is well funded through grants from the National Science Foundation and European Commission.

To provide better context around the data used in research and better expose data for reuse, journals are now adding "data papers". These are specifically designed for publishing articles about the data, and sharing it. This benefits both researchers and readers. Researchers receive credit for data they have collected or produced, and readers get more context about the data.

@@

Whilst useful, these requirements do not tell the author how to actually structure the data and folders for reuse. Instead, they provide ideas on what they should include. This is a useful step towards improving data reuse, but it _lacks some minimal structure that allows a researcher to have a predictable way to access and interpret the data_.

@@

The data are free to use and publicly accessible, with the interface to the data being designed to cover a wide range of skills. For example, the marvin service to streamline access and visualisation of MaNGA data [@marvin-service], through to raw data access (see Figure \@ref(fig:sdss-raw-data-screenshot)).

@@

- Metadata in the machine readable standard, SCHEMA
- Data in multiple formats (e.g., .csv, .fits, and database)
- Previously released data available in archives
- Entire datasets available for download
- Raw data before processed into database is also available

@@

1. **Decide whether publishing is appropriate for your data** 
It is critical to publish your data if they are the basis for a peer-reviewed research publication. To be broadly useful, your data must be deposited in a permanent archive, to ensure that it does not disappear from a ephemeral location such as your university website. It should also contain useful metadata so that someone outside the project team can understand what the data, and variables, mean. However, this level of effort is not always critical for all data science efforts. For small-scale projects such as ones where one might generate simulated datasets, this level of curation is highly unnecessary. Here making data available in a transient location like GitHub or a website is sufficient.

@@

10. **Stick with simple data formats** to ensure long-term usefulness of datasets in a way that is not tied to transient technologies (such as ever changing spreadsheet standards), store the data as plain text (e.g., CSV) where possible. This can take the form of comma, or tab separated files in most cases. This will ensure transparency, and future proof your data.

@@

Over the years researchers have made data available in a multitude of unreliable ways, including offering downloads from university websites, FTP repositories, and password protected servers, all of which are prone to bit rot. In rare cases where the data are deposited in a permanent archive, insufficient metadata and missing data processing scripts have made it harder and more time intensive to map these to code.  Our goal here is to describe a variety of ways a researcher can ship data, analogous to how easy it has become to push code to services like GitHub. The choice of technology and method all depends on the nature of the data, and its size and complexity. Once the issue of access is solved, we also discuss how much metadata to include in order to make the data useful. Ultimately there is a tradeoff to consider on a case by case basis. Long-term studies with complex protocols require structured metadata, while more standardized datasets such as those coming from sensors require far less effort.
